---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Load the required libraries

```{r, warning = FALSE, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

Download and unzip dataset

```{r, warning = FALSE, message = FALSE}
url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile<-paste0(getwd(),"/activity_monitoring_data.zip")
download.file(url,destfile)
unzip("activity_monitoring_data.zip")
```

Read the activity dataset csv into R

```{r, warning = FALSE, message = FALSE}
activity_data <- read.csv(paste0(getwd(),"/activity.csv"))
```


## What is mean total number of steps taken per day?

```{r, warning = FALSE, message = FALSE}
daily_steps <- activity_data %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps, na.rm = TRUE))

mean_steps <- daily_steps %>%
  summarise(daily_steps = mean(total_steps, na.rm = TRUE))

median_steps <- daily_steps %>%
  summarise(daily_steps = median(total_steps, na.rm = TRUE))
```

The mean of the total number of steps taken per day is `r mean_steps` and the median of the total number of steps taken per day is 
`r median_steps`.

Ignoring NAs, the mean total number of steps taken per day is `r mean(daily_steps$total_steps, na.rm = TRUE)`.


```{r, warning = FALSE, message = FALSE}
qplot(daily_steps$total_steps, geom="histogram")
```

## What is the average daily activity pattern?

```{r, warning = FALSE, message = FALSE}
interval_steps <- activity_data %>%
  group_by(interval) %>%
  summarise(average_steps = mean(steps, na.rm = TRUE))

interval_steps <- as.data.frame(interval_steps)

ggplot(interval_steps, aes(interval, average_steps)) + geom_line() +
  xlab("Interval") + ylab("Average Steps")

```

The interval with the maximum average steps across all the days is interval `r interval_steps[which.max(interval_steps[,2]),1]`.

## Imputing missing values

activity_data[complete.cases(activity_data), ]

```{r, warning = FALSE, message = FALSE}
NAs <- nrow(((activity_data[rowSums(is.na(activity_data)), ])))
```

The total number of rows with NA in the activity dataset is `r NAs`.

The following code fills the missing values in the dataset by replacing NAs with the mean steps observed in the affected interval. Otherwise the dataset is identical.

```{r, warning = FALSE, message = FALSE}
activity_data <- activity_data %>%
  group_by(interval) %>% 
  mutate(steps=ifelse(is.na(steps),mean(steps,na.rm=TRUE),steps)) %>%
  ungroup()
```

```{r, warning = FALSE, message = FALSE}
daily_steps <- activity_data %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps, na.rm = FALSE))

mean_steps <- daily_steps %>%
  summarise(daily_steps = mean(total_steps, na.rm = FALSE))

median_steps <- daily_steps %>%
  summarise(daily_steps = median(total_steps, na.rm = FALSE))

qplot(daily_steps$total_steps, geom="histogram") 
```

With missing values imputed, the mean number of steps is now `r mean_steps` and the median number of steps is now `r median_steps`, roughly 10% higher compared to when missing values were ignored. In addition, the mean is now equal to the median.


### Are there differences in activity patterns between weekdays and weekends?

```{r, warning = FALSE, message = FALSE}
activity_data <- activity_data %>%
  mutate(weekday = weekdays(as.POSIXct(date))) %>%
  mutate(weekday = case_when(weekday %in% c("Saturday", "Sunday") ~ "weekend",
                             weekday %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")  ~ "weekday")) %>%
  mutate(weekday = as.factor(weekday)) 

weekday_steps <- activity_data %>%
  filter(weekday == "weekday") %>%
  group_by(interval) %>%
  summarise(average_steps = mean(steps))

weekend_steps <- activity_data %>%
  filter(weekday == "weekend") %>%
  group_by(interval) %>%
  summarise(average_steps = mean(steps))

layout(matrix(c(1,2), nrow = 2, ncol = 1, byrow = TRUE))
plot(weekday_steps$interval, weekday_steps$average_steps, type='l', xlab = '', ylab= "Average steps", col="blue", lwd=2 ,main="weekday")
plot(weekend_steps$interval, weekend_steps$average_steps, type='l', xlab = "Interval", ylab= "Average steps", col="blue", lwd=2,main="weekend")

```

Plotting weekday and weekend activity data seperately shows that activity is distributed more evenly throughout the day on weekends unlike the weekday that sees a clear peak around 8-9 AM. Activity appears to be overall higher on weekends than on weekdays, making for a pronounced difference in activity patterns between weekdays and weekends.
